---
title: ""
author: ""
date: ""
output:
  pdf_document:
    latex_engine: xelatex  
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(dplyr)
library(sf)
library(ggplot2)
library(corrplot)
library(lubridate)
library(systemfonts)
library(GGally)
library(knitr)
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(pheatmap)
library(reshape2)
library(spdep)      # For Moran's I and spatial weights
library(spatstat)   # For Ripley's K-function
library(viridis)
library(ggpubr)
library(ggcorrplot)

grep("Arial", system_fonts()$family, value = TRUE)
```

# Table of Contents

\newpage

```{r, echo = FALSE, warning = FALSE, include = FALSE}
data_cleaned <- read.csv("cleaned_water_data.csv")
coordinates <- read.csv("coordinates.csv")
```


```{r, warning = FALSE, message = FALSE, fig.align = 'center'}
station_counts <- data_cleaned %>%
  group_by(Water_Object) %>%
  summarise(Number_of_Stations = n())%>%
  arrange(desc(Number_of_Stations))
```

```{r}
top_bottom_counts <- bind_rows(
  head(station_counts, 5), 
  tail(station_counts, 5)  
)

top_bottom_counts
```

```{r}
ggplot(station_counts, aes(x = reorder(Water_Object, -Number_of_Stations), y = Number_of_Stations)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black", width = 0.7) +
  labs(
    title = "Number of Sampling Stations per Water Object",
    subtitle = "Visualizing the distribution of stations across various water objects",
    x = "Water Object",
    y = "Number of Stations"
  ) +
  theme_classic(base_family = "Arial Unicode MS") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, family = "Arial Unicode MS"), 
    plot.subtitle = element_text(size = 12, hjust = 0.5, family = "Arial Unicode MS"),
    axis.text.x = element_text(angle = 90, hjust = 1, family = "Arial Unicode MS"),
  )
```
```{r}
arm_sf <- st_read('Armenia_Marzes.shp', quiet = TRUE) 
```

```{r}
arm_df <- arm_sf %>% group_by(MarzID) %>%
  st_cast("POINT")
arm_df = data.frame(Marz = arm_df$Name_Eng, st_coordinates(arm_df))
```


```{r}
ggplot(arm_df, aes(x = X, y = Y, group = Marz)) +
  geom_polygon(fill = 'white', color = 'black') + coord_fixed()+ theme_void()
```

```{r}
coordinates_clean <- coordinates %>%
  mutate(
    Longitude = as.numeric(Longitude),
    Latitude = as.numeric(Latitude)
  ) %>%
  filter(!is.na(Longitude) & !is.na(Latitude))  # Remove rows with NA in coordinates

# Check for duplicates and remove them
coordinates_clean <- coordinates_clean %>%
  distinct(Longitude, Latitude, .keep_all = TRUE)
```
```{r}
colSums(is.na(coordinates_clean))
```











Hypothesis 1: There is a spatial correlation between water quality indicators (e.g., BOD5, COD) and the geographic location (X, Y coordinates) of the sampling stations.







**Exploratory Data Analysis (EDA)**

```{r}
# Univariate Analysis - Histogram of Dissolved Oxygen
ggplot(data_cleaned, aes(x = Dissolved_Oxygen)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Dissolved Oxygen", x = "Dissolved Oxygen (mg/L)", y = "Frequency")
```


**Outlier Detection**

**Time Series Analysis**


**Correlation Analysis**

```{r}
# Select numeric columns
numeric_cols <- data_cleaned %>%
  select(where(is.numeric))

# Compute correlation matrix
cor_matrix <- cor(numeric_cols, use = "complete.obs")
pairs(numeric_cols[, 1:8], main = "Scatterplot Matrix for Selected Variables")
```

**Principal Component Analysis (PCA)**

```{r}
numeric_data <- data_cleaned %>%
  select(where(is.numeric))
```

```{r}
numeric_data <- numeric_data %>%
  select(where(~ all(!is.na(.)))) 
```

```{r}
numeric_data_scaled <- scale(numeric_data)
```

```{r}
data_pca = prcomp(x = numeric_data_scaled, center = T, scale. = T)
summary(data_pca)
```

```{r}
pheatmap(numeric_data_scaled, cluster_rows = T, cluster_cols = T)
```

```{r}
fviz_screeplot(data_pca, addlabels = T)
```

```{r}
fviz_pca_biplot(data_pca, axes = c(1, 2), geom = c("point", "text"), label = "var")
```


```{r}
fviz_pca_var(data_pca, col.var = "black") 
```

**Clustering**

```{r}
# Use scaled numeric data from the PCA section for clustering
data_for_clustering <- numeric_data_scaled

set.seed(123)  # For reproducibility
fviz_nbclust(data_for_clustering, FUNcluster = factoextra::hcut,
method = "wss")

```

```{r}
fviz_nbclust(data_for_clustering, kmeans, method = "wss") +
labs(title = "K-means Scree Plot")

```
```{r}
# Apply K-means clustering
set.seed(123)
optimal_k <- 3  # Based on the elbow plot
kmeans_result <- kmeans(data_for_clustering, centers = optimal_k, nstart = 25)

# Add cluster labels to the original data
data_cleaned$Cluster <- as.factor(kmeans_result$cluster)

# Create a scatter plot using two selected columns
ggplot(data_cleaned, aes(x = Dissolved_Oxygen, y = BOD5, color = Cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "K-means Clustering Visualization",
    x = "Dissolved Oxygen (mg/L)",
    y = "BOD5 (mg/L)",
    color = "Cluster"
  ) +
  scale_color_brewer(palette = "Set2")

```


```{r}
# Step 3: Hierarchical Clustering
# Compute dissimilarity matrix
dist_matrix <- dist(data_for_clustering, method = "euclidean")

# Perform hierarchical clustering
hclust_result <- hclust(dist_matrix, method = "ward.D2")

# Plot the dendrogram
plot(hclust_result, labels = FALSE, hang = -1,
     main = "Dendrogram of Hierarchical Clustering",
     xlab = "Samples",
     ylab = "Height")

# Cut the dendrogram into clusters
hclust_clusters <- cutree(hclust_result, k = 3)

# Add hierarchical clustering labels to the original data
data_cleaned$HCluster <- as.factor(hclust_clusters)

```



**Geospatial Analysis**


```{r}
# Filter data for "Sevan Lake" and "Debed river basin"
filtered_data <- data_cleaned %>%
  filter(River_Basin %in% c("Sevan Lake", "Debed river basin", "Metsamor river basin"))

# Create a visually appealing boxplot
ggplot(filtered_data, aes(x = River_Basin, y = Dissolved_Oxygen, fill = River_Basin)) +
  geom_boxplot(outlier.color = "red", outlier.size = 2) + # Highlight outliers
  labs(
    title = "Dissolved Oxygen by River Basin",
    subtitle = "Comparing Sevan Lake and Debed River Basin",
    x = "River Basin",
    y = "Dissolved Oxygen (mg/L)"
  ) +
  scale_fill_manual(values = c("Sevan Lake" = "#1f77b4", "Debed river basin" = "#ff7f0e", "Metsamor river basin"='green')) + # Custom colors
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "none" # Hide legend as it's clear from labels
  )


```
#Correlation Heatmap

```{r}
# Filter only key numeric columns for correlation analysis
selected_columns <- c("Dissolved_Oxygen", "BOD5", "COD", "NH4", "NO3", "Zn", "Cu", "Pb", "Hardness", "TDS", "Suspended_Solids")
filtered_data <- data[selected_columns]

# Compute the correlation matrix
cor_matrix_filtered <- cor(filtered_data, use = "complete.obs")

ggcorrplot(
  cor_matrix_filtered,
  lab = TRUE, # Show correlation values
  type = "lower", # Show only the lower triangle
  colors = c("blue", "white", "red"), # Color scale
  title = "Clean Correlation Heatmap of Selected Water Quality Parameters",
  outline.color = "black",
  tl.cex = 0.8 # Adjust text size for better readability
)
```
The variables selected for the correlation heatmap are chosen based on their importance in understanding water quality and their interrelation in assessing environmental and ecological conditions. Here’s an explanation of why these specific variables are included and what they reveal:

Strong Positive Correlations:

BOD5 and COD (0.88): These are both measures of organic pollution, indicating they often increase together in polluted environments.
COD and NH4 (0.77): Indicates that chemical pollution (COD) is often associated with higher ammonium levels, likely due to organic matter decomposition.
TDS and Suspended Solids (0.66): Suggests that higher dissolved solids are often found alongside particulate matter, reflecting general water pollution.
Moderate Correlations:

TDS and Metals (e.g., Zn 0.44): Dissolved solids include metal ions, showing a direct connection between TDS and metal pollution levels.
Weak or Negative Correlations:

DO and NH4 (-0.13): Dissolved oxygen decreases slightly as ammonium levels increase, reflecting oxygen consumption during ammonium nitrification.
DO and BOD5 (-0.32): Dissolved oxygen decreases with higher organic pollution, consistent with oxygen demand from organic matter decomposition.
Key Insights:
Organic and chemical pollution indicators (BOD5, COD, NH4) are closely interrelated, highlighting sources like sewage or industrial discharge.
Suspended solids and TDS are linked to overall water pollution, potentially including heavy metals.
Dissolved oxygen is negatively impacted by pollution, affecting aquatic life.



#Trend analysis for different metals

```{r}
# Trend analysis for different metals
metals <- c("Zn", "Cu", "Cr", "As", "Cd", "Pb")
metal_data <- data_cleaned %>% select(Date, all_of(metals)) %>% melt(id.vars = "Date")

ggplot(metal_data, aes(x = Date, y = value, color = variable)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Trend Analysis of Metal Concentrations Over Time", x = "Date", y = "Concentration") +
  facet_wrap(~variable, scales = "free_y")
  labs(title = "Histogram of Suspended Solids", x = "Suspended Solids", y = "Frequency")
```
#Hypothesis 6: Spatial Clustering of Stations
##Hypothesis: Water monitoring stations are spatially clustered rather than evenly distributed across the country.

```{r}
# Convert Longitude and Latitude to numeric
coordinates_clean <- coordinates %>%
  mutate(
    Longitude = as.numeric(Longitude),
    Latitude = as.numeric(Latitude)
  ) %>%
  filter(!is.na(Longitude) & !is.na(Latitude))  # Remove rows with NA in coordinates

# Check for duplicates and remove them
coordinates_clean <- coordinates_clean %>%
  distinct(Longitude, Latitude, .keep_all = TRUE)

# Verify the ranges are valid
print(range(coordinates_clean$Longitude, na.rm = TRUE))
print(range(coordinates_clean$Latitude, na.rm = TRUE))

```


```{r}
# Load the shapefile of Armenia
armenia_map <- st_read("Armenia_Marzes.shp")

# Load the coordinates dataset
coordinates <- read.csv("coordinates.csv")

# Step 1: Clean the data
# Convert Longitude and Latitude to numeric, coercing invalid values to NA
coordinates$Longitude <- as.numeric(coordinates$Longitude)
coordinates$Latitude <- as.numeric(coordinates$Latitude)

# Filter out rows with NA in the River_Basin_Area column
coordinates_clean <- coordinates_clean %>%
  filter(!is.na(River_Basin_Area))

# Create an sf object for the monitoring stations
stations_sf <- st_as_sf(coordinates_clean, coords = c("Longitude", "Latitude"), crs = 4326)

# Plot monitoring stations on Armenia's map without NA
ggplot() +
  geom_sf(data = armenia_map, fill = "lightgrey", color = "black") +
  geom_sf(data = stations_sf, aes(color = River_Basin_Area), size = 2) +
  scale_color_viridis_d() +
  labs(title = "Monitoring Stations Across Armenia", color = "River Basin Area") +
  theme_minimal()


```
```{r}
sum(is.na(coordinates$Longitude))
sum(is.na(coordinates$Latitude))
range(coordinates_clean$Longitude)  # Should return a valid numeric range
range(coordinates_clean$Latitude)   # Should return a valid numeric range
# Identify non-numeric Longitude and Latitude entries
non_numeric_long <- coordinates_clean[is.na(as.numeric(coordinates_clean$Longitude)), ]
non_numeric_lat <- coordinates_clean[is.na(as.numeric(coordinates_clean$Latitude)), ]

# Print problematic rows
print(non_numeric_long)
print(non_numeric_lat)
```

```{r}
# Transform Armenia map and stations to the same CRS
armenia_map <- st_transform(armenia_map, crs = 32638)  # EPSG: 32638 is UTM zone 38N
stations_sf <- st_as_sf(coordinates_clean, coords = c("Longitude", "Latitude"), crs = 4326)
stations_sf <- st_transform(stations_sf, crs = st_crs(armenia_map))
```

```{r}
# Ensure Moran's I input is correct
coords_matrix <- st_coordinates(stations_sf)  # Extract numeric coordinates
nb <- knn2nb(knearneigh(coords_matrix, k = 5))  # Create nearest neighbors
listw <- nb2listw(nb, style = "W")  # Create spatial weights

# Compute Moran's I (for example, on Longitude or Latitude)
moran_test <- moran.test(coordinates_clean$Longitude, listw)
print(moran_test)
```
###Why Moran's I Test is Used
Moran's I is a statistical measure of spatial autocorrelation. It quantifies the degree to which a spatial variable is correlated with itself across space. It helps answer whether nearby spatial units (in this case, monitoring stations) tend to have similar values (clustering) or dissimilar values (dispersion).

###Purpose in Spatial Analysis:
Test whether water monitoring stations are spatially clustered (positive autocorrelation), randomly distributed (no autocorrelation), or dispersed (negative autocorrelation).

###Analysis of the Results
Moran I Statistic:

Value: 0.9679 (close to 1).
Interpretation: Strong positive spatial autocorrelation, indicating that water monitoring stations are spatially clustered.
p-value:

Value: < 2.2e-16 (very low).
Interpretation: The result is statistically significant, rejecting the null hypothesis of random distribution.
Alternative Hypothesis:

Greater: Indicates testing for positive spatial autocorrelation (clustering).


```{r}
# Load required libraries
library(sp)
library(maptools)

# Transform Armenia map to SpatialPolygons if it is not already
armenia_sp <- as(armenia_map, "Spatial")

# Extract bounding box from the Armenia map
bounding_box <- bbox(armenia_sp)

# Convert the bounding box to a window object (owin)
bounding_window <- as.owin(list(
  xrange = bounding_box[1, ],
  yrange = bounding_box[2, ]
))

```

```{r}
# Convert stations_sf to a ppp object for Ripley’s K-function
coords_ppp <- as.ppp(st_coordinates(stations_sf), W = bounding_window)

# Compute Ripley's K-function
library(spatstat)
k_function <- Kest(coords_ppp, correction = "border")

# Plot the results
plot(k_function, main = "Ripley's K-function Analysis")
```
Why Ripley's K-Function is Used
Ripley's K-function is a spatial statistics tool used to analyze the distribution of spatial point patterns. It helps identify whether the points (e.g., monitoring stations) are clustered, randomly distributed, or regularly spaced across different spatial scales.

Purpose in Spatial Analysis:
Understand clustering patterns over different distances (r values).
Evaluate if clustering or dispersion exists at specific spatial scales, beyond what Moran's I provides.

Components of the Plot
Black Line (Kbord(r)):

Represents the observed K-function, showing the cumulative number of points within a distance r for the observed data.
Higher values indicate clustering over increasing distances.
Red Dashed Line (Kpois(r)):

Represents the expected K-function under the null hypothesis of Complete Spatial Randomness (CSR).
Comparison:

If the black line lies above the red dashed line, it indicates clustering.
If the black line lies below the red dashed line, it indicates dispersion.

Analysis of the Results
Clustering Indicated:

The observed K-function (Kbord(r)) is consistently above the expected K-function (Kpois(r)).
This indicates strong clustering of monitoring stations at all spatial scales analyzed (r values ranging from 0 to 70,000 meters).
Increasing Clustering Over Distance:

As the distance (r) increases, the gap between the observed K-function and the expected K-function widens.
This suggests that clustering becomes more pronounced at larger spatial scales, potentially reflecting regional grouping of monitoring stations.


```{r}

# Plot kernel density estimation with enhancements
ggplot() +
  # Add Armenia's map as a base layer
  geom_sf(data = armenia_map, fill = "lightgrey", color = "black", alpha = 0.5) +
  # Add kernel density with enhanced color gradient
  stat_density2d(
    data = as.data.frame(coords_matrix),
    aes(x = X, y = Y, fill = ..level..),
    geom = "polygon", contour = TRUE,
    alpha = 0.8
  ) +
  # Customize the fill color scale
  scale_fill_viridis_c(option = "C", name = "Density Level") +
  # Add labels and improve title formatting
  labs(
    title = "Kernel Density of Monitoring Stations in Armenia",
    subtitle = "Visualizing spatial clustering of water monitoring stations",
    x = "Longitude",
    y = "Latitude"
  ) +
  # Adjust theme for clarity
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    legend.position = "right",
    legend.key.height = unit(1, "cm"),
    legend.key.width = unit(0.5, "cm")
  )
```
Spatial Distribution:

The plot highlights clusters of monitoring stations in specific areas.
Areas with high density (yellow regions) indicate regions of priority or interest, such as major water bodies(Sevan) or industrial areas.


```{r}
# Enhanced scatter plot with colors
ggscatter(
  data_cleaned,
  x = "Suspended_Solids",
  y = "Dissolved_Oxygen",
  add = "reg.line",
  conf.int = TRUE, 
  cor.coef = TRUE,
  cor.method = "pearson", 
  xlab = "Total Suspended Solids (mg/L)",
  ylab = "Dissolved Oxygen (mg/L)",
  title = "Relationship Between Total Suspended Solids and Dissolved Oxygen",
  color = "Suspended_Solids"
) +
  scale_color_viridis_c(
    option = "C",                          
    name = "TSS (mg/L)"                    
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(
      face = "bold", size = 14, hjust = 0.5, color = "#2c3e50"
    ), 
    axis.title = element_text(
      size = 14, face = "bold", color = "#34495e"
    ),
    axis.text = element_text(size = 12, color = "#34495e"),
    panel.grid.major = element_line(color = "#ecf0f1"), 
    panel.grid.minor = element_blank(), 
    legend.position = "right", 
    legend.title = element_text(face = "bold", color = "#2c3e50"),
    legend.key = element_rect(fill = "white")
  )


```
The weak negative correlation (R=−0.041) suggests that, generally, higher concentrations of Total Suspended Solids are associated with slightly lower Dissolved Oxygen levels.

Significance:
Despite the weak correlation, the significant p-value (p=0.034) indicates that the relationship is statistically meaningful, but the effect size is small. Other factors likely contribute to variations in DO.

```{r}
coordinates$Station_Number <- as.character(coordinates$Station_Number)
data_cleaned$Station_Number <- as.character(data_cleaned$Station_Number)

```


#Hypothesis 2: Dominance of Specific Water Objects
##Hypothesis: Certain water objects (e.g., major rivers or lakes) account for the majority of water monitoring stations due to their importance in water management.
```{r}
# Step 1: Group by Water_Object and count the number of stations
water_object_summary <- data_cleaned %>%
  group_by(Water_Object) %>%
  summarise(Number_of_Stations = n()) %>%
  arrange(desc(Number_of_Stations))  # Sort by descending number of stations

# Step 2: Limit to Top N Water Objects for readability (e.g., top 20)
top_water_objects <- water_object_summary %>%
  top_n(20, Number_of_Stations)

# Step 3: Create a beautiful bar plot with further adjustments for x-axis labels
ggplot(top_water_objects, aes(x = reorder(Water_Object, -Number_of_Stations), y = Number_of_Stations, fill = Number_of_Stations)) +
  geom_bar(stat = "identity", show.legend = FALSE, alpha = 0.9) +  # Bar chart
  scale_fill_viridis_c(option = "C") +  # Color gradient for bars
  labs(
    title = "Dominance of Specific Water Objects",
    subtitle = "Top 20 water objects by the number of monitoring stations",
    x = "Water Object",
    y = "Number of Monitoring Stations"
  ) +
  theme_minimal(base_size = 14) +  # Modern minimal theme
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#2c3e50"),
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "#34495e"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),  # Increase angle and adjust alignment
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    panel.grid.major = element_line(color = "#ecf0f1"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 20, 20, 50)  # Increase left margin for better label spacing
  )

```

