---
title: ""
author: ""
date: ""
always_allow_html: true
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)
library(readxl)
library(dplyr)
library(sf)
library(ggplot2)
library(corrplot)
library(lubridate)
library(systemfonts)
library(networkD3)
library(GGally)
library(tidyr)
library(knitr)
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(pheatmap)
library(reshape2)
library(spdep)      
library(spatstat)  
library(viridis)
library(ggpubr)
library(ggcorrplot)
```

```{r, echo = FALSE, warning = FALSE, include = FALSE}
data_cleaned <- read.csv("cleaned_water_data.csv")
coordinates <- read.csv("coordinates.csv")
```

\begin{titlepage}
    \centering
    {\Huge \textbf{Data Visualization} \par}
    \vspace{1cm}  % Adds 1 cm of vertical space
    {\Huge \textbf{DS116 / CS343} \par}
    \vspace{3cm}  % Adds 1 cm of vertical space
    {\huge \textbf{Final Project} \par}
    \vspace{0cm}  % Adds 1 cm of vertical space
    {\LARGE \textbf{Water Data} \par}
    \vspace{5cm}  % Adds 2 cm of vertical space
    {\large Elen Ghalechyan\\Hripsime Soghomonyan\\Marieta Khachatryan\\Marina Melkonyan \par}
    \vspace{5cm}  % Adds 1 cm of vertical space
    \vfill
    {\large American University of Armenia\\Yerevan, Armenia\\December 6, 2024\par}
\end{titlepage}

\section*{\Huge Overview}

Water quality is one of the essential factors according to which the nature of water can be researched. Water quality directly impacts drinking water safety and agricultural productivity. Nowadays, many countries face a problem connected to the lack of drinking water. At the same time, Armenia is valued for its water resources how people state, especially foreign tourists; whenever you walk around the streets, you will come to a fountain from which water is constantly flowing, and the most exciting part is that you can use it and drink. This idea is highly connected with ecosystems and public health in Armenia; thus, understanding water quality is essential for addressing problems such as pollution, managing resources, and promoting environmental well-being. 

If we consider this report from the perspective of the benefit of stakeholders, then we can state that a lot of them will benefit. The government can use this report to establish policies regulating pollutants and protecting water resources. Also, water structures can be controlled; thus, we can shift to environmental agencies that can monitor patterns of water quality degradation to react swiftly and mitigate risks. 

\section*{\Huge Research Methodology}
The research was done with the help of 2 main datasets. The first and primary dataset contains detailed measurements of water quality parameters for the monitoring stations, such as dissolved oxygen, ammonium levels, chemical oxygen demand (COD), suspended solids, water hardness level, etc. The second dataset includes the geographic coordinates of the monitoring stations, enabling spatial analysis of water quality trends within and across river basins.

The datasets were collected in Armenian, requiring translation into English for easy analysis. Additional preprocessing steps included standardizing date and time formats, merging the datasets using station identifiers, and addressing missing values through imputation or exclusion. 

The R programming language was used for the analysis. Additionally, we developed an interactive dashboard using the "Shiny" package. 

\section*{\Huge Literature Review}
Water quality in Armenia is vital because industrial, agricultural, and urban activities have increasingly impacted water quality. Two key reports provide valuable insights into Armenia's water resources: The Araks River Basin and the World Bank report on water resource management challenges. The report first looks at the Araks River Basin, including the most critical issues: pollution from heavy metals, turbidity, and untreated discharges. Arsenic and chromium are often elevated in tributaries of the Hrazdan and Sevjur from industrial and mining activities. These findings underscore the significance of pollution control targeting and improved wastewater treatment to conserve the basin's ecological and public health. The World Bank report introduces broader issues of water governance, infrastructure, and use, which are also raised in the context of Armenia's water resource management. However, this confirms the fragmentation of water management systems through multiple institutional responsibilities, which undermines efficient policymaking. The report recommends an integrated approach to water management, combining policy reforms, infrastructure investments, and data-driven monitoring to overcome Armenia's growing pressures on its water resources. These reports give a complete picture of Armenia's water quality problems. The World Bank report, while providing strategic recommendations for systemic improvements in water governance and infrastructure, focuses on broader determinants of pollution. For further details, refer to the https://www.researchgate.net/publication/225259203_Assessing_the_Quality_of_Water_of_the_Araks_Basin_Rivers_in_Armeniaand and  https://documents1.worldbank.org/curated/en/099062424121038546/pdf/P1700301619bb50718c6b1c554e0879f1e.pdf.

\section*{\Huge Data}

The dataset being analyzed during the analysis contains vital water quality measurements collected from various monitoring stations across multiple water bodies. Each row represents a specific sampling event, with columns providing detailed information on physical, chemical, and biological parameters measured during that event. The dataset includes variables such as Dissolved Oxygen (mg/L), Biochemical Oxygen Demand (BOD5) (mg/L), and Chemical Oxygen Demand (COD) (mg/L), which are critical indicators of organic pollution and water health. Additionally, it includes nutrient concentrations (e.g., Ammonium, Nitrate, and Phosphate), metal levels (e.g., Zinc, Lead, and Copper), and general information like Sampling Date, Station Location, and Water Object.
The temporal coverage spans from the earliest recorded sample date (`r min(data_cleaned$Date)`) to the latest (`r max(data_cleaned$Date)`), enabling trend analysis over time. Numeric variables provide quantitative measurements essential for understanding variations and correlations within the dataset.
This structure ensures a comprehensive dataset, allowing in-depth analysis of water quality trends, pollution sources, and potential ecological impacts. 

\section*{\Huge Hypothesis and Analysis}

\section*{Analysis: Distribution of Sampling Stations Across Water Objects}


\textbf{Null Hypothesis (H0)}: Sampling stations are equally distributed across different water objects.\newline
\textbf{Alternative Hypothesis (H1)}: Sampling stations are not equally distributed across different water objects.


First, we examine the distribution of sampling stations across several water objects to understand the representativeness of the dataset. This allows us to identify if some water objects are over or under represented and subsequently can influence future analyses. We can get some initial insights into the balance of the dataset by visualizing the distribution and then summarizing the top and bottom water objects by the number of stations.

```{r, include=FALSE}
station_counts <- data_cleaned %>%
  group_by(Water_Object) %>%
  summarise(Number_of_Stations = n())%>%
  arrange(desc(Number_of_Stations))
total_stations <- station_counts %>% summarise(Total_Stations = sum(Number_of_Stations))
```

```{r, include=FALSE}
top_bottom_counts <- bind_rows(
  head(station_counts, 5), 
  tail(station_counts, 5)  
)
# top_bottom_counts
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(station_counts, aes(x = reorder(Water_Object, -Number_of_Stations), y = Number_of_Stations)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black", width = 0.7) +
  labs(
    title = "Number of Sampling Stations per Water Object",
    subtitle = "Visualizing the distribution of stations across various water objects",
    x = "Water Object",
    y = "Number of Stations"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

There is a significant variation between the number of stations throughout the water objects. For a better understanding, 5 water objects with the most number of stations are filtered.  

```{r, echo=FALSE, message=FALSE, warning=FALSE}
top_5 <- head(station_counts, 5)

ggplot(top_5, aes(x = reorder(Water_Object, -Number_of_Stations), y = Number_of_Stations)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black", width = 0.7) +
  geom_label(aes(label = Number_of_Stations), vjust = 0.5, size = 4) + 
  labs(
    title = "Top 5 Water Objects by Sampling Stations",
    x = "Water Object",
    y = "Number of Stations"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```
The water objects with the most number of stations are Great and Small Sevans, Hrazdan, Pambak and Qasakh. The analysis reveals that the top five water objects with the highest number of sampling stations align with expectations, as these represent the primary water sources for Armenia. These water objects are known for their significant contribution to the country's water supply system and play a critical role in meeting both domestic and industrial water demands. Their prominence in the dataset highlights their importance within the broader context of Armenia’s hydrological resources.

```{r, include=FALSE}
station_counts_1 <- data_cleaned %>%
  group_by(Water_Object) %>%
  summarise(Number_of_Stations = n())
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(station_counts_1, aes(y = Number_of_Stations)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  theme_classic() +
  labs(
    title = "Statistical Distribution of Station Counts Across Water Objects",
    y = "Number of Stations",
    x = ""
  ) +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        axis.text.x = element_blank())
```
From the boxplot, it can be observed that the majority of water objects have relatively few monitoring stations, as indicated by the clustering within the box. The median number of stations is low, suggesting that half of the water objects have fewer than this value. However, there are a number of significant outliers above the upper whisker. The overall distribution is right-skewed, reflecting a monitoring strategy focused on larger or more significant water objects. While this prioritization is logical, the low representation of many water objects suggests that smaller or less prominent areas may not be receiving adequate attention. This imbalance could leave certain ecological or human health risks unaddressed.

To better understand the placement and distribution of the water objects and their sampling stations, it is essential to visualize them on a map. This spatial perspective will provide insights into how the stations are geographically distributed and whether the current monitoring strategy effectively covers critical water resources.

```{r, include=FALSE}
armenia_map <- st_read("Armenia_Marzes.shp")
stations_sf <- st_as_sf(coordinates, coords = c("Longitude", "Latitude"), crs = 4326)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot() +
  geom_sf(data = armenia_map, fill = "lightgrey", color = "black") +
  geom_sf(data = stations_sf, aes(color = Water_Object), size = 2) +
  scale_color_viridis_d() +
  labs(title = "Monitoring Stations Across Armenia") +
  theme_classic() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text.x = element_text(hjust = 1),
    legend.position = "none" # Removes the legend
  )
```

The spatial distribution of monitoring stations across Armenia, visualized on the map, provides a comprehensive understanding of their geographical placement in relation to key water bodies. This complements the earlier statistical findings, where prominent water objects such as Great Sevan, Small Sevan, and major rivers like Hrazdan were identified as having the highest number of sampling stations. The map highlights a strategic focus on these critical resources, which play an essential role in Armenia’s water supply and ecosystem health. However, it also underscores potential gaps in coverage, particularly in less monitored regions, which aligns with the boxplot analysis suggesting underrepresentation of smaller or less prominent water bodies.

\section*{Analysis: Relationships Between Water Objects and Key Pollutants}

\textbf{Null Hypothesis (H0)}: The distribution and levels of key pollutants (Dissolved Oxygen, COD, NH4, Pb, and SO4) are consistent across all selected water objects.\newline
\textbf{Alternative Hypothesis (H1)}: The distribution and levels of key pollutants vary significantly across the selected water objects.


The aim of this analysis is to visualize and understand the differences in pollution patterns across selected water objects by focusing on their associations with key contaminants. To achieve this, the top 5 water objects with the highest number of monitoring stations and the bottom 5 water objects with the least number of monitoring stations have been selected. This selection highlights the variability between heavily monitored and less monitored water objects, providing a balanced view of contamination dynamics.

The pollutants included in this analysis—Dissolved Oxygen (DO), Chemical Oxygen Demand (COD), Ammonium (NH4), Lead (Pb), and Sulfate (SO4)—were chosen for their environmental significance and diverse origins. DO is a critical indicator of water health, with low levels often signaling pollution or ecosystem stress. COD reflects organic pollution levels and is a direct measure of water contamination. NH4 is linked to nutrient pollution from agricultural runoff or wastewater discharge, contributing to eutrophication. Pb, a toxic heavy metal, is often associated with industrial or urban pollution, posing significant ecological and public health risks. SO4 serves as an indicator of industrial or mining contamination, affecting water usability and aquatic habitats.

Logically, we expect the Sankey diagram to reveal distinct patterns of contamination. Heavily monitored water objects might exhibit diverse pollution profiles due to their ecological or strategic importance, while less monitored objects may show narrower or less consistent associations. We also anticipate variations in pollutant levels, with industrial regions showing higher heavy metal concentrations and agricultural areas demonstrating elevated nutrient pollution. This analysis will provide a detailed understanding of these differences, guiding targeted interventions for water quality improvement.\newline


```{r, include=FALSE}
# Rank water objects by the number of stations (or another metric)
top_5_objects <- station_counts %>% top_n(5, Number_of_Stations)
least_5_objects <- station_counts %>% top_n(-5, Number_of_Stations)

# Combine top and least water objects
selected_objects <- bind_rows(top_5_objects, least_5_objects)

# Identify key pollutant columns
pollutants <- c("Dissolved_Oxygen", "COD", "NH4", "Pb", "SO4") 

# Filter the dataset for selected water objects
contaminant_data <- data_cleaned %>%
  filter(Water_Object %in% selected_objects$Water_Object) %>%  
  select(Water_Object, all_of(pollutants)) %>% 
  pivot_longer(
    cols = all_of(pollutants),
    names_to = "Contaminant_Type",
    values_to = "Contaminant_Level"
  ) %>%
  group_by(Water_Object, Contaminant_Type) %>%
  summarise(Avg_Level = mean(Contaminant_Level, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Avg_Level))

```

```{r, include=FALSE}
nodes <- data.frame(
  name = c(
    unique(contaminant_data$Water_Object), 
    unique(contaminant_data$Contaminant_Type) 
  )
)


links <- contaminant_data %>%
  mutate(
    source = match(Water_Object, nodes$name) - 1,  
    target = match(Contaminant_Type, nodes$name) - 1 
  ) %>%
  select(source, target, Avg_Level) %>%
  rename(value = Avg_Level) 
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Create the Sankey Diagram
sankeyNetwork(
  Links = links,        
  Nodes = nodes,             
  Source = "source",        
  Target = "target",    
  Value = "value",    
  NodeID = "name",           
  units = "Average Level",     
  fontSize = 14,        
  nodeWidth = 30,
  colourScale = 'd3.scaleOrdinal(["#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854", "#ffd92f"])'
)
```
Among the pollutants, Chemical Oxygen Demand (COD) has largest flows associated with the Lake Sevan (Great Sevan), indicating that there exists organic pollution in excess, probably due to a discharge or runoff resulting from industrial activity. In addition, COD is highly correlated with other water bodies including Hrazdan and Pambak; however, flow thicknesses are heterogeneous further implying different pollution sources. Relationship of sulfate (SO4) with Small Sevan and Arpi Lake Reservoir suggests such activities as industrial and mining related activities or natural geochemical processes. Herher and Kasakh are ammonium (NH4), which are with farming or without effective sewage discharge, especially in zones near farming. Less prominent lead (Pb) connections appear to represent heavy metal contamination at localized water objects, possibly induced by industrial or urban sources nearby. Finally, dissolved oxygen (DO) levels linked to multiple water bodies display which areas are stressed ecologically and reduced DO frequently indicates poor water quality and eutrophication. This diagram serves to illustrate how different degrees and types of pollution affect different water objects, conveying the spatial complexity and dynamics of contamination that are useful to understand to plan for targeted water management.


\section*{Analysis: Temporal Trends in Water Quality Components}

\textbf{Null Hypothesis (H0)}: There are no significant temporal trends in the concentrations of Dissolved Oxygen (DO), Chemical Oxygen Demand (COD), Ammonium (NH4), Lead (Pb), and Sulfate (SO4) across Lake Sevan, Hrazdan River, Pambak River, and other selected water objects.\newline
\textbf{Alternative Hypothesis (H1)}: There are significant temporal trends in the concentrations of Dissolved Oxygen (DO), Chemical Oxygen Demand (COD), Ammonium (NH4), Lead (Pb), and Sulfate (SO4) across Lake Sevan, Hrazdan River, Pambak River, and other selected water objects.

The long-term sustainability of water resources in Armenia depends on understanding how key water quality parameters evolve over time. This analysis focuses on the temporal trends of pollutants across selected water bodies, including Lake Sevan (Great Sevan and Small Sevan), Hrazdan River, and Pambak River, among others. The pollutants under investigation—Dissolved Oxygen (DO), Chemical Oxygen Demand (COD), Ammonium (NH4), Lead (Pb), and Sulfate (SO4)—are critical indicators of water health. Temporal trends in these components reflect the impact of anthropogenic activities such as agricultural runoff, industrial discharges, and urban pollution. By identifying patterns or significant changes over time, this analysis aims to detect potential ecological degradation, evaluate seasonal dynamics, and provide insights into pollution control efforts.






















**Analysis**: Analysis of Water Quality Components and Limitations

While the map provides a visual representation of the distribution of sampling stations across various water objects, the next step is to shift focus to the water quality components that were analyzed.These components, such as dissolved oxygen, chemical oxygen demand (COD), and various chemical contaminants, offer crucial insights into the quality of the water. The following analysis will explore the concentrations and trends observed for these components, highlighting any significant variations or patterns that emerge from the data. It is important to note that the number of sources included and samples taken in the data is relatively limited and, additionally, some components were adjusted through multiplication by specific factors (?as advised by our professor). These adjustments, while necessary for analysis, may not perfectly align with actual conditions. Consequently, the interpretations of the data might lead to misconceptions or overgeneralizations. 


```{r}
# Univariate Analysis - Histogram of Dissolved Oxygen
ggplot(data_cleaned, aes(x = Dissolved_Oxygen)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Dissolved Oxygen", 
    x = "Dissolved Oxygen (mg/L)",
    y = "Frequency") + 
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))  

```
The histogram illustrates the distribution of dissolved oxygen (DO) levels in water samples, revealing several key patterns. The distribution of DO is highly right-skewed, with the majority of values concentrated between 0 and 15 mg/L.Within this range a significant increase can be observed between 5 and 10 mg/L, which is considered to be the optimal DO level. However, there is a notable spike at or near zero indicating the presence of hypoxic conditions in certain water samples. Hypoxic conditions occur when DO levels in water drop below 2 mg/L, which is insufficient to support most aquatic life. These conditions are generally harmful to the health of aquatic ecosystems. This highlights the fact that there are stations which are either close to or in "dead zones". 



#Correlation Heatmap

```{r}
# Filter only key numeric columns for correlation analysis
selected_columns <- c("Dissolved_Oxygen", "BOD5", "COD", "NH4", "NO3", "Zn", "Cu", "Pb", "Hardness", "TDS", "Suspended_Solids")
filtered_data1 <- data_cleaned[selected_columns]

# Compute the correlation matrix
cor_matrix_filtered <- cor(filtered_data1, use = "complete.obs")

# Plot the correlation heatmap
ggcorrplot(
  cor_matrix_filtered,
  lab = TRUE,                    # Show correlation values
  type = "lower",                # Show only the lower triangle
  colors = c("blue", "white", "red"),  # Color scale
  title = "Correlation Heatmap of Water Quality Parameters",
  outline.color = "black",       # Outline the cells for clarity
  lab_size = 3.5,                # Adjust the size of the correlation labels
  tl.cex = 10                    # Adjust the size of the axis labels
) +
  theme(
    plot.title = element_text(
      face = "bold", size = 16, hjust = 0.5, color = "#2c3e50"
    ),                            # Bold and centered title
    axis.text.x = element_text(
      angle = 45, hjust = 1, size = 10
    ),                            # Rotate and adjust font size of x-axis labels
    axis.text.y = element_text(size = 10) # Adjust font size of y-axis labels
  )
```
This heatmap paints a clear picture of how different water quality factors are connected and influence one another. For instance, we see that BOD5 (Biochemical Oxygen Demand) and COD (Chemical Oxygen Demand) are strongly linked. This makes sense because both measure the impact of organic pollution in water — when there's more organic matter breaking down, both of these measures increase. Similarly, COD and ammonium (NH4) show a close relationship, likely because organic waste and agricultural runoff contribute to higher ammonium levels, which in turn raise chemical oxygen demand.

Another important insight comes from the connection between Suspended Solids and Total Dissolved Solids (TDS). As these solids increase, it's often a sign of general pollution in the water. Metals like zinc (Zn), copper (Cu), and lead (Pb) show varying levels of connection to dissolved solids, which suggests that metals may also be part of this pollution but come from different sources like industrial runoff or mining.

We can see that the heatmap reveals a concerning trend for Dissolved Oxygen (DO), which is critical for aquatic life. As pollutants like BOD5, COD, and NH4 increase, oxygen levels tend to drop. This is because these pollutants demand more oxygen as they break down, leaving less available for fish and other aquatic organisms. Low oxygen levels, especially in heavily polluted areas, can harm ecosystems, making these findings crucial for environmental action.

**Hypothesis (H₀)**: The levels of nitrate (NO3), phosphate (PO4), and ammonium (NH4) are evenly distributed across Armenia’s river basins and show no significant seasonal variation.
**Alternative Hypothesis (H₁)**: Nutrient pollution varies significantly between river basins and seasons due to agricultural, industrial, and seasonal runoff differences.

```{r}
data_cleaned$Season <- cut(as.numeric(format(as.Date(data_cleaned$Date), "%m")),
                           breaks = c(0, 3, 6, 9, 12),
                           labels = c("Winter", "Spring", "Summer", "Fall"))

nutrient_summary <- data_cleaned %>%
  group_by(River_Basin, Season) %>%
  summarise(
    Avg_NO3 = mean(NO3, na.rm = TRUE),
    Avg_NH4 = mean(NH4, na.rm = TRUE),
    Avg_PO4 = mean(PO4, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = starts_with("Avg"), 
               names_to = "Nutrient", 
               values_to = "Concentration")


# Temporal trends visualization
temporal_trends <- data_cleaned %>%
  group_by(Date) %>%
  summarise(
    Avg_NO3 = mean(NO3, na.rm = TRUE),
    Avg_NH4 = mean(NH4, na.rm = TRUE),
    Avg_PO4 = mean(PO4, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = starts_with("Avg"), 
               names_to = "Nutrient", 
               values_to = "Concentration")

ggplot(temporal_trends, aes(x = as.Date(Date), y = Concentration, color = Nutrient)) +
  geom_line(size = 1.2) +
  scale_color_viridis_d() +
  labs(
    title = "Temporal Trends in Nutrient Pollution",
    x = "Date",
    y = "Concentration (mg/L)",
    color = "Nutrient"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    legend.position = "right"
  )

```

Nutrient pollution, including nitrates, phosphates, and ammonium, is crucial for understanding water quality changes over time. This pollution often stems from agricultural practices and urban runoff, with seasonal variations, such as spring snowmelt or summer farming activities, further exacerbating its impact. By analyzing these temporal and spatial trends, it becomes clear how pollution hotspots emerge and how these nutrient levels fluctuate. The findings strongly demonstrate the influence of both human activities and natural seasonal factors on nutrient dynamics.

Ammonium (NH4) exhibited significant variability, with noticeable spikes in late 2023 and early 2024, likely linked to agricultural runoff during planting and harvesting seasons, industrial discharges, or untreated sewage. This irregular behavior highlights the combined impact of human activities and seasonal events. In contrast, nitrate (NO3) levels remained relatively stable, suggesting a steady input from consistent sources, such as farming. Phosphate (PO4) levels, on the other hand, were consistently low, indicating effective regulatory controls or limited sources of phosphate pollution in Armenia.

The observed patterns—ammonium's variability, nitrate's steadiness, and phosphate's low levels—provide strong evidence that nutrient pollution is shaped by both anthropogenic and seasonal factors. Therefore, the null hypothesis, which assumes no temporal variability in nutrient pollution, is rejected. The alternative hypothesis, asserting the influence of human and seasonal drivers, is accepted.s

\section*{\Huge Conclusion}

\section*{\Huge References}

Nalbandyan, A., Avagyan, A., & Sargsyan, G. (2011). Assessing the Quality of Water of the Araks Basin Rivers in Armenia. Environmental Science and Pollution Research. Retrieved from https://www.researchgate.net/publication/225259203_Assessing_the_Quality_of_Water_of_the_Araks_Basin_Rivers_in_Armenia

The World Bank. (2024). Armenia Water Sector Overview: Strengthening the Institutional and Governance Frameworks. World Bank Group. Retrieved from https://documents1.worldbank.org/curated/en/099062424121038546/pdf/P1700301619bb50718c6b1c554e0879f1e.pdf

