---
title: "DS 116 - Data Visualization"
subtitle: "Water Analysis"
author: "Elen Ghalechyan,Hripsime Soghomonyan,Marieta Khachatryan,Marina Melkonyan Section A"
institute: "American University of Armenia" 
date: "2024-12-06"
fontsize: 10pt
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "dolphin"
    fonttheme: "structurebold"
    pandoc_args: '--pdf-engine=xelatex'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = F,  collapse = TRUE, warning=F, message = F, out.width = '80%', 
                      size = 'footnotesize')
options(scipen = 999)
library(readxl)
library(dplyr)
library(sf)
library(ggplot2)
library(corrplot)
library(lubridate)
library(systemfonts)
library(networkD3)
library(GGally)
library(tidyr)
library(knitr)
library(ggcorrplot)
library(FactoMineR)
library(RColorBrewer)
library(factoextra)
library(pheatmap)
library(reshape2)
library(spdep)      
library(spatstat)  
library(viridis)
library(ggpubr)
library(ggcorrplot)
```

```{r, echo = F}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "footnotesize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\footnotesize"), x)
})
```

```{r, echo = FALSE, warning = FALSE, include = FALSE}
data_cleaned <- read.csv("cleaned_water_data.csv")
coordinates <- read.csv("coordinates.csv")
```

## Overview

- Armenia’s water quality is crucial for drinking water safety and agricultural productivity.
- Objectives:
  - Analyze distribution of water sampling stations.
  - Assess relationships between pollutants and water objects.
  - Explore eutrophication risks and propose interventions.
  
## Research Methodology

- **Datasets**:
  - Primary: Water quality parameters (DO, COD, NH4, metals, etc.).
  - Secondary: Geographic coordinates of sampling stations.
- **Preprocessing**:
  - Translation, merging, and handling missing values.
- **Analysis conducted in R**, with Shiny dashboard for visualization.

## Distribution of Sampling Stations

We analyzed the number of monitoring stations across various water objects to understand the representativeness of the dataset.

```{r, include=FALSE}
station_counts <- data_cleaned %>%
  group_by(Water_Object) %>%
  summarise(Number_of_Stations = n())%>%
  arrange(desc(Number_of_Stations))
total_stations <- station_counts %>% summarise(Total_Stations = sum(Number_of_Stations))
```

```{r, include=FALSE}
top_bottom_counts <- bind_rows(
  head(station_counts, 5), 
  tail(station_counts, 5)  
)
# top_bottom_counts
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(station_counts, aes(x = reorder(Water_Object, -Number_of_Stations), y = Number_of_Stations)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black", width = 0.7) +
  labs(
    title = "Number of Sampling Stations per Water Object",
    subtitle = "Visualizing the distribution of stations across various water objects",
    x = "Water Object",
    y = "Number of Stations"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5), 
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
top_5 <- head(station_counts, 5)

ggplot(top_5, aes(x = reorder(Water_Object, -Number_of_Stations), y = Number_of_Stations)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black", width = 0.7) +
  geom_label(aes(label = Number_of_Stations), vjust = 0.5, size = 4) + 
  labs(
    title = "Top 5 Water Objects by Sampling Stations",
    x = "Water Object",
    y = "Number of Stations"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5), 
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

```{r, include=FALSE}
station_counts_1 <- data_cleaned %>%
  group_by(Water_Object) %>%
  summarise(Number_of_Stations = n())
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(station_counts_1, aes(y = Number_of_Stations)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  theme_classic() +
  labs(
    title = "Statistical Distribution of Station Counts Across Water Objects",
    y = "Number of Stations",
    x = ""
  ) +
  theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
        axis.text.x = element_blank())
```

```{r, include=FALSE}
armenia_map <- st_read("Armenia_Marzes.shp")
stations_sf <- st_as_sf(coordinates, coords = c("Longitude", "Latitude"), crs = 4326)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot() +
  geom_sf(data = armenia_map, fill = "lightgrey", color = "black") +
  geom_sf(data = stations_sf, aes(color = Water_Object), size = 2) +
  scale_color_viridis_d() +
  labs(title = "Monitoring Stations Across Armenia") +
  theme_classic() +
  theme(
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5), 
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text.x = element_blank(),    
    axis.text.y = element_blank(),         
    axis.ticks = element_blank(),
    legend.position = "none"
  )
```

```{r, include=FALSE}
top_5_objects <- station_counts %>% top_n(5, Number_of_Stations)
least_5_objects <- station_counts %>% top_n(-5, Number_of_Stations)

selected_objects <- bind_rows(top_5_objects, least_5_objects)

pollutants <- c("Dissolved_Oxygen", "COD", "NH4", "Pb", "SO4") 

contaminant_data <- data_cleaned %>%
  filter(Water_Object %in% selected_objects$Water_Object) %>%  
  select(Water_Object, all_of(pollutants)) %>% 
  pivot_longer(
    cols = all_of(pollutants),
    names_to = "Contaminant_Type",
    values_to = "Contaminant_Level"
  ) %>%
  group_by(Water_Object, Contaminant_Type) %>%
  summarise(Avg_Level = mean(Contaminant_Level, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Avg_Level))

```

```{r, include=FALSE}
nodes <- data.frame(
  name = c(
    unique(contaminant_data$Water_Object), 
    unique(contaminant_data$Contaminant_Type) 
  )
)


links <- contaminant_data %>%
  mutate(
    source = match(Water_Object, nodes$name) - 1,  
    target = match(Contaminant_Type, nodes$name) - 1 
  ) %>%
  select(source, target, Avg_Level) %>%
  rename(value = Avg_Level) 
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
# sankeyNetwork(
#   Links = links,
#   Nodes = nodes,
#   Source = "source",
#   Target = "target",
#   Value = "value",
#   NodeID = "name",
#   units = "Average Level",
#   fontSize = 14,
#   nodeWidth = 30,
#   colourScale = 'd3.scaleOrdinal(["#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854", "#ffd92f"])'
# )
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data_cleaned, aes(x = Dissolved_Oxygen)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Dissolved Oxygen", 
    x = "Dissolved Oxygen (mg/L)",
    y = "Frequency") + 
  theme_classic() +
  theme(
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5))  
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
selected_columns <- c("Dissolved_Oxygen", "BOD5", "COD", "NH4", "NO3", "Zn", "Cu", "Pb", "Hardness", "TDS", "Suspended_Solids")
filtered_data1 <- data_cleaned[selected_columns]

cor_matrix_filtered <- cor(filtered_data1, use = "complete.obs")

ggcorrplot(
  cor_matrix_filtered,
  lab = TRUE,                    
  type = "lower",               
  colors = c("blue", "white", "red"),  
  title = "Correlation Heatmap of Water Quality Parameters",
  outline.color = "black",   
  lab_size = 2,             
  tl.cex = 10                  
) +
  labs(
    x = NULL,                          
    y = NULL                               
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(
      face = "bold", size = 13, hjust = 0.5, color = "#2c3e50"
    ),                          
    axis.text.x = element_text(
      angle = 45, hjust = 1, size = 10
    ),                          
    axis.text.y = element_text(size = 10) 
  )
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Compute N:P ratio
data_cleaned$NP_Ratio <- data_cleaned$Total_Inorganic_N / data_cleaned$PO4

# Add a risk flag for easier analysis
data_cleaned$Eutrophication_Risk <- ifelse(data_cleaned$NP_Ratio < 16, "Phosphorus-Limited", "Nitrogen-Limited")

# Scatter plot visualizing the relationship between inorganic Nitrogen and Phosphate
ggplot(data_cleaned, aes(x = PO4, y = Total_Inorganic_N, color = Eutrophication_Risk)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_manual(values = c("Phosphorus-Limited" = "steelblue", "Nitrogen-Limited" = "darkred")) +
  labs(
    title = "Scatterplot of N:P Ratio for Eutrophication Risk",
    x = "Phosphate (PO4) Concentration (mg/L)",
    y = "Total Inorganic Nitrogen (mg/L)",
    color = "Eutrophication Risk Type"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5))
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
merged_data <- stations_sf %>%
  left_join(data_cleaned %>% select(Station_Number, Eutrophication_Risk), by = "Station_Number") %>%
  filter(!is.na(Eutrophication_Risk))

ggplot() +
  geom_sf(data = armenia_map, fill = "lightgrey", color = "black") +  
  geom_sf(data = merged_data, aes(color = Eutrophication_Risk), size = 2) +  
  scale_color_manual(values = c("Phosphorus-Limited" = "steelblue", "Nitrogen-Limited" = "darkred")) +
  labs(title = "Eutrophication Risk Across Monitoring Stations",
       color = "Eutrophication Risk Type") +
  theme_classic() + 
  theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5),axis.text.x = element_blank(),
        axis.text.y = element_blank(),axis.ticks = element_blank())
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data_cleaned$Season <- cut(as.numeric(format(as.Date(data_cleaned$Date), "%m")),
                           breaks = c(0, 3, 6, 9, 12),
                           labels = c("Winter", "Spring", "Summer", "Fall"))

nutrient_summary <- data_cleaned %>%
  group_by(River_Basin, Season) %>%
  summarise(
    Avg_NO3 = mean(NO3, na.rm = TRUE),
    Avg_NH4 = mean(NH4, na.rm = TRUE),
    Avg_PO4 = mean(PO4, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = starts_with("Avg"), 
               names_to = "Nutrient", 
               values_to = "Concentration")


# Temporal trends visualization
temporal_trends <- data_cleaned %>%
  group_by(Date) %>%
  summarise(
    Avg_NO3 = mean(NO3, na.rm = TRUE),
    Avg_NH4 = mean(NH4, na.rm = TRUE),
    Avg_PO4 = mean(PO4, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = starts_with("Avg"), 
               names_to = "Nutrient", 
               values_to = "Concentration")

ggplot(temporal_trends, aes(x = as.Date(Date), y = Concentration, color = Nutrient)) +
  geom_line(size = 1) +
  scale_color_viridis_d() +
  labs(
    title = "Temporal Trends in Nutrient Pollution",
    x = "Date",
    y = "Concentration (mg/L)",
    color = "Nutrient"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    legend.position = "right"
  )

```